Finite Time Bounds for Temporal Difference Learning with Function Approximation: Bugs in state-of-the-art results

### 2017-07-30 issues.tex
pass on issues

The first equality of their display states (in equivalent form) that 
\begin{align*}
A^\top A = \norm{\phi}^2 ( \phi \phi^\top - 2\beta \phi (\phi')^\top +\beta^2 \phi' (\phi')^\top )\,.
\end{align*}
We see that while this is closer to the correct result, here the mistake is that $-2\beta \phi  (\phi')^\top$ is replacing $-\beta(\phi (\phi')^\top + \phi' \phi^\top)$.
\begin{comment}
We also have a hard time following the rest of the proof of this lemma.
In fact, we have a simple proof for a result similar to stated in this lemma.
This works by observing that one needs to 
bound the Lipschitz factor of $\theta \mapsto \bar T_n \dots \bar T_{t+1} ( (I-\gamma_t A_t)\theta + b_t )$, where
$A_t = \phi_t (\phi_t -\beta \phi_t')^\top$, $b_t = r_{i_t} \phi_t$ and
 $\bar T_t(\theta) = (I-\gamma_t \bar A_T) \theta + \bar b_T$ (we assume that the stepsize sequence is deterministic).
 Clearly, this Lipschitz factor can be bounded by $\prod_{s=t+1}^n \norm{I-\gamma_s \bar A_T}_2 \norm{I-\gamma_t A_t}_2$
 (the composition of Lipschitz maps is Lipschitz with a factor that is the product of the individual maps' Lipschitz factors). 
Invoking (A3) then indeed gives some version of Lemma 7.
By making an appropriate assumption (similar to those used in our paper), 
$\prod_{s=t+1}^n \norm{I-\gamma_s \bar A_T}_2 $ can be controlled.
\end{comment}
\subsection{Bugs in the paper by  \cite{lstdicmla}}
In page 14 the expression below "A is a possibly random matrix..." is not entirely justified (personal communication with one of the authors).

\section{Bounded Linear Space Assumption}
\subsection{Bugs in the paper by \citet{flstda}}
\citet{flstda} assume that the parameter vector stays such that the value function $\Phi \theta$ (with their notation) will be bounded in $L^\infty$-norm (see assumption \textbf{$(A3)$} of \cite{flstd} and \cite{flstda}). This assumption is critical in establishing Lemma~7 (see pages 15 and 16, \cite{flstda}).
We suspect the following shortcomings with assumption \textbf{$(A3)$}
\begin{itemize}
\item The assumption is stated in a somewhat sloppy fashion. We take the authors meant to say that $\sup_n \norm{\Phi \theta_n}_\infty<+\infty$ holds almost surely. This seems like a strong assumption! Would not one of the main question be whether this assumption even holds?
% We see in our paper that for some cases, the iterates can blow up if the step-size is inappropriately selected!
\item One possibility would be to modify the algorithm by adding a projection step to guarantee boundedness.
It is still unclear  whether this alone would ensure convergence of the error to zero. In any case, the expected error bound analysis is invalidated if a projection step is present (basically, the algebraic identities will all fail to hold).

%For the sake of an argument let us also additionally assume that the iterates are bounded by a projection step (which  \citet{flstda} don't mention explicitly). Then would the recursion unfold the same way as it does in Equation (26) of \cite{flstda} (or Equation (13) of \cite{flstd})? How are we to ensure decay/contraction (as shown in Equation (27) of \cite{flstda}) at each step (since projection might be non-contracting)?
%\item The recursion eventually (see Equation (26) of \cite{flstda}) has only \emph{additive noise} and not \emph{multiplicative noise} as analyzed in our submission. The additive noise is captured in the form of the $H_\beta^2$ term in Equation (28) of \cite{flstd}. This reduction to additive noise is possible due to \textbf{$(A3)$}.
\end{itemize}
\subsection{Bugs in the paper by \citet{lstdicmla}}
\citet{gugan} mention that (citing a personal communication with \citet{lstdicmla}) \cite{lstdicmla} assume implicitly the projection step in all the high probability bounds in \cite{lstdicmla}. While this implicit projection in itself does not affect the high probability bound proofs directly, it is important to note that the set to which the iterates needs to be projected has to contain the TD(0) solution, a condition that cannot be guaranteed.

\section{Relation between Covariance Matrix and $\bar{A}_T$ matrix}
\citet{flstda} assume positive definiteness of \textbf{$(A4)$} covariance matrix $\frac{1}{T}\Phi^\top_T \Phi_T$. However, unlike regression problems, in reinforcement learning problems what appears in the recursion (see Equation (6)) is not the covariance matrix, but a different matrix $\bar{A}_T=\frac{1}{T}\sum_{i=1}^T \phi(s_i)(\phi(s_i)-\beta\phi(s_i'))^\top$ defined in pages 2, 4 (below Equation (5)), 8 and 16 of \cite{flstda}. Usually, without a sampling assumption known as the `on-policy' case (see \cite{gtd} for a discussion on `on-policy' vs `off-policy') the eigenvalues of $\bar{A}_T$ cannot be guaranteed to have all positive real parts. While \citet{flstda} mention the `on-policy' sampling in the introduction, there is no explicit sampling assumption in the list of assumption. In fact, we doubt that the proposed algorithm will converge without extra assumption (as discussed above).
\if0
This is worrisome for the following reasons
\begin{itemize}
\item \textbf{$(A4)$} deals only with the covariance matrix and it is not clear how that alone is sufficient to say that $\bar{A}_n-(1-\beta)\mu$ is positive definite (see below Equation (26) page 16 of \cite{flstda}; the problem mentioned earlier).
\item For the sake of the argument, let us suppose that $(1-\beta)\mu$ is the smallest eigenvalue of $\bar{A}_n$. However, the problem here is that $\bar{A}_n$ is not symmetric in general and it is typical in reinforcement learning applications to say a non-symmetric matrix $M$ is positive definite if $x^\top M x>0$ for all real valued vectors $x$. While it follows from this notion of positive definiteness of $M$ that all the eigenvalues of $M$ have positive real parts, there could still be eigenvalues with complex conjugate parts. Hence to assume that the smallest eigenvalue of $\bar{A}_n$ to be $(1-\beta)\mu$ might be problematic (because there could be other complex eigenvalues with smaller magnitude). While it might be true in practice that the smallest eigenvalue of $\bar{A}_n$, is greater than $(1-\beta)\mu$ to our knowledge there is no proof for this fact in literature.
\end{itemize}
\fi
\section{Blow Up of the Bound}
We would like to note that the rate expression in Corollary~4 of \cite{flstda} (or Corollary~2 of \cite{flstd}) contains a constant $C$. The authors do mention that the sampling error blows up as $\alpha \rightarrow 1$.
However, it also looks like that even the constant $C=\sum_{n=1}^\infty \exp(-\mu cn^{1-\alpha})$ will blow up as $\alpha\rightarrow 1$, in which case it is not clear whether a $1/\sqrt{n}$ rate is achieved (even if we disregard all the other problems).

\section{Doubt about the Effectiveness of the Algorithms}
In Corollary~4 of \cite{flstda} (or Corollary~2 of \cite{flstd}), we learn that the value of $c$ governing the stepsize of the primary update must be in a small range (it must be between $1.33$ and $2$). This means, that effectively, the stepsize $\gamma_n$ behaves as $1/n^\alpha$ ($c$ has very little effect). At least when $\alpha=1$, we know that stepsizes like this make the bias decrease slowly and averaging remains ineffective. This seems to be at odds with the suggestion after this result that $\alpha\to 1$ is a desirable choice. In fact, we would be inclined to choose $\alpha=1/2$, i.e., its lowest value. This is because then the bias is decreased relatively quickly, while the variance will be controlled by the additional averaging. However, given all the problems with this it remains to be seen whether this is indeed a reasonable choice and under exactly what conditions this is reasonable.
\bibliographystyle{plainnat}
\bibliography{issuesref}
%\newpage
%\input{linal}

>>>>

The first equality of their display states (in equivalent form) that 
\begin{align*}
A^\top A = \norm{\phi}^2 ( \phi \phi^\top - 2\beta \phi (\phi')^\top +\beta^2 \phi' (\phi')^\top )\,.
\end{align*}
We see that while this is closer to the correct result, here the mistake is that $-2\beta \phi  (\phi')^\top$ is replacing $-\beta(\phi (\phi')^\top + \phi' \phi^\top)$.
\begin{comment}
We also have a hard time following the rest of the proof of this lemma.
In fact, we have a simple proof for a result similar to stated in this lemma.
This works by observing that one needs to 
bound the Lipschitz factor of $\theta \mapsto \bar T_n \dots \bar T_{t+1} ( (I-\gamma_t A_t)\theta + b_t )$, where
$A_t = \phi_t (\phi_t -\beta \phi_t')^\top$, $b_t = r_{i_t} \phi_t$ and
 $\bar T_t(\theta) = (I-\gamma_t \bar A_T) \theta + \bar b_T$ (we assume that the stepsize sequence is deterministic).
 Clearly, this Lipschitz factor can be bounded by $\prod_{s=t+1}^n \norm{I-\gamma_s \bar A_T}_2 \norm{I-\gamma_t A_t}_2$
 (the composition of Lipschitz maps is Lipschitz with a factor that is the product of the individual maps' Lipschitz factors). 
Invoking (A3) then indeed gives some version of Lemma 7.
By making an appropriate assumption (similar to those used in our paper), 
$\prod_{s=t+1}^n \norm{I-\gamma_s \bar A_T}_2 $ can be controlled.
\end{comment}
\subsection{Bugs in the paper by  \cite{lstdicmla}}
In page 14 the expression below ``$A$ is a possibly random matrix...'' is not justified (personal communication with one of the authors confirmed this).  In particular, the claim here is that if $A$ is a random matrix with $\norm{A}_2\le C$ with $C$ a deterministic constant then for any $\theta$ deterministic vector,
\begin{align*}
\EE{ \theta^\top A^\top \EE{ \epsilon_n | \mathcal{F}_n } A \theta \,|\, s_0 } \le C^2 \theta^\top \EE{ \epsilon_n|s_0 } \theta\,.
\end{align*}
Recall that here $\epsilon_n$ is a matrix of appropriate dimensions, the ``mixing-error term''
and is actually $\mathcal{F}_n$ measurable ($\epsilon_n = \EE{ a_n | \mathcal{F}_n } - \mathbb{E}_{\Psi}[a_n]$). 
When the Markov chain is started from its stationary state 
(which is not ruled out by the conditions of the theorem under question), $\EE{ \epsilon_n|s_0 } = 0$.
If the above inequality was true, we would get 
\begin{align*}
\EE{ A^\top \epsilon_n  A  \,|\, s_0 } = 0\,.
\end{align*}
However, letting $B = \EE{\epsilon_n|\mathcal{F}_n}$ and, for example,
$A = C B/\norm{ B }_2$, we have 
\begin{align*}
A^\top \EE{ \epsilon_n | \mathcal{F}_n } A  = \frac{C^2}{\norm{ B }_2^2} B^\top B B
\end{align*}
and it is easy to construct examples where the expectation of this is nonzero.

\section{Issues with the Setup}
\subsection{Boundedness of iterates: \citet{flstda}}
\citet{flstda} assume that the parameter vector stays such that the value function $\Phi \theta$  will be bounded in $L^\infty$-norm (see assumption (A3) of \cite{flstd} and \cite{flstda}). This assumption is critical in establishing Lemma~7 (see pages 15 and 16, \cite{flstda}), in an argument that is similar to the proof of McDiarmid's inequality.
We suspect the following shortcomings with assumption (A3):
\begin{itemize}
\item The assumption is stated in a somewhat sloppy fashion. We take the authors meant to say that $\sup_n \norm{\Phi \theta_n}_\infty<+\infty$ holds almost surely. This seems like a strong assumption:  ensuring this will most likely further restrict the step-size sequences that can be used. The step-size sequences that give the best rate under (A3) may include step-size sequences which in fact lead to $\Prob{\limsup_{n\to\infty}\norm{\theta_n}=\infty}>0$. Without proving that this is not the case, the results of the paper have limited utility.
% We see in our paper that for some cases, the iterates can blow up if the step-size is inappropriately selected!
\item One possibility would be to modify the algorithm by adding a projection step to guarantee boundedness.
It is still unclear  whether this alone would ensure convergence of the error to zero. In any case, the expected error bound analysis is invalidated if a projection step is present (basically, the algebraic identities will all fail to hold) and a new proof will be required.

%For the sake of an argument let us also additionally assume that the iterates are bounded by a projection step (which  \citet{flstda} don't mention explicitly). Then would the recursion unfold the same way as it does in Equation (26) of \cite{flstda} (or Equation (13) of \cite{flstd})? How are we to ensure decay/contraction (as shown in Equation (27) of \cite{flstda}) at each step (since projection might be non-contracting)?
%\item The recursion eventually (see Equation (26) of \cite{flstda}) has only \emph{additive noise} and not \emph{multiplicative noise} as analyzed in our submission. The additive noise is captured in the form of the $H_\beta^2$ term in Equation (28) of \cite{flstd}. This reduction to additive noise is possible due to \textbf{$(A3)$}.
\end{itemize}
\subsection{Boundedness of iterates:  \citet{lstdicmla}}
\citet{gugan} mention that (citing a personal communication with \citet{lstdicmla}) 
that \cite{lstdicmla} assume implicitly a projection step in all the high probability bounds. 
While this implicit projection in itself does not affect the high probability bound proofs directly, 
the algebraic steps are invalidated.
Furthermore, the set that the iterates are projected to should contain the TD(0) solution.
How to ensure this (without knowing $A,b$) remains to be seen.

\subsection{Relation between Covariance Matrix and $\bar{A}_T$ matrix}
\citet{flstda} assume positive definiteness of \textbf{$(A4)$} covariance matrix $\frac{1}{T}\Phi^\top_T \Phi_T$. However, unlike regression problems, in reinforcement learning problems what appears in the recursion (see Equation (6)) is not the covariance matrix, but a different matrix $\bar{A}_T=\frac{1}{T}\sum_{i=1}^T \phi(s_i)(\phi(s_i)-\beta\phi(s_i'))^\top$ defined in pages 2, 4 (below Equation (5)), 8 and 16 of \cite{flstda}. Usually, without a sampling assumption known as the `on-policy' case (see \cite{gtd} for a discussion on `on-policy' vs `off-policy') the eigenvalues of $\bar{A}_T$ cannot be guaranteed to have all positive real parts. While \citet{flstda} mention the `on-policy' sampling in the introduction, there is no explicit sampling assumption in the list of assumption. In fact, we doubt that the proposed algorithm will converge without extra assumption (as discussed above).
\if0
This is worrisome for the following reasons
\begin{itemize}
\item \textbf{$(A4)$} deals only with the covariance matrix and it is not clear how that alone is sufficient to say that $\bar{A}_n-(1-\beta)\mu$ is positive definite (see below Equation (26) page 16 of \cite{flstda}; the problem mentioned earlier).
\item For the sake of the argument, let us suppose that $(1-\beta)\mu$ is the smallest eigenvalue of $\bar{A}_n$. However, the problem here is that $\bar{A}_n$ is not symmetric in general and it is typical in reinforcement learning applications to say a non-symmetric matrix $M$ is positive definite if $x^\top M x>0$ for all real valued vectors $x$. While it follows from this notion of positive definiteness of $M$ that all the eigenvalues of $M$ have positive real parts, there could still be eigenvalues with complex conjugate parts. Hence to assume that the smallest eigenvalue of $\bar{A}_n$ to be $(1-\beta)\mu$ might be problematic (because there could be other complex eigenvalues with smaller magnitude). While it might be true in practice that the smallest eigenvalue of $\bar{A}_n$, is greater than $(1-\beta)\mu$ to our knowledge there is no proof for this fact in literature.
\end{itemize}
\fi
\subsection{Blow Up of the Bound}
We would like to note that the rate expression in Corollary~4 of \cite{flstda} (or Corollary~2 of \cite{flstd}) contains a constant $C$. The authors do mention that the sampling error (a.k.a. variance) blows up as $\alpha \rightarrow 1$.
However, it also looks like that even the constant $C=\sum_{n=1}^\infty \exp(-\mu cn^{1-\alpha})$ (appearing in the bound of the bias) will blow up as $\alpha\rightarrow 1$, in which case it seems that the claim that the $1/\sqrt{n}$ rate can be achieved in the limit will not hold.

\subsection{Doubt about the Effectiveness of the Algorithms}
In Corollary~4 of \cite{flstda} (or Corollary~2 of \cite{flstd}), we learn that the value of $c$ governing the stepsize of the primary update must be in a small range (it must be between $1.33$ and $2$). This means, that effectively, the stepsize $\gamma_n$ behaves as $1/n^\alpha$ ($c$ has very little effect). At least when $\alpha=1$, we know that stepsizes like this make the bias decrease slowly and averaging remains ineffective. This seems to be at odds with the suggestion after this result that $\alpha\to 1$ is a desirable choice. In fact, we would be inclined to choose $\alpha=1/2$, i.e., its lowest value. This is because then the bias is decreased relatively quickly, while the variance will be controlled by the additional averaging. However, given all the problems with this, it remains to be seen whether this is indeed a reasonable choice and under exactly what conditions this is reasonable.
\bibliographystyle{plainnat}
\bibliography{issuesref}
%\newpage
%\input{linal}
